---
title: "Exercise 1 --- Basic Maps and Graphs in R"
subtitle: "BSTA 374 · 12 points · Due Thursday, September 10, 2026, 11:59 pm"
author: "Mamun Ur Rashid Pranta · Lehigh University"
date: last-modified
format:
  html:
    theme: cosmo
    toc: true
    toc-depth: 2
    toc-location: left
    toc-title: "On this page"
    number-sections: false
    code-copy: true
    highlight-style: github
    embed-resources: true
  pdf:
    papersize: letter
    geometry:
      - margin=1in
    toc: false
    colorlinks: true
    code-overflow: wrap
prefer-html: true
execute:
  eval: false
  echo: true
---

## What this exercise is

Your first graded work in this course. It covers the first two weeks: reading
spatial data with `sf`, understanding what a coordinate reference system is and
moving between them, and producing honest choropleth and point maps with both
`ggplot2` and `tmap`.

**No data to download.** Everything runs on the North Carolina SIDS data that
ships inside the `sf` package, so your analysis will reproduce on my machine
exactly as it does on yours --- which is the point.

Each part ends with **starting code**. Copy it, fill in the blanks marked
`______`, and adapt it. It is scaffolding, not a solution: the mechanical lines
are given so that your time goes into the parts that require judgement.

::: {.callout-important}
## Submission
Submit **one file** to Course Site:

- `LASTNAME_FI_Exercise1.qmd` --- the source

Example: `DELMELLE_EM_Exercise1.qmd`. Put your name in the document header.

**Due Thursday, September 10, 2026 at 11:59 pm.** Late work loses 20% per day
unless you arrange something with me in advance.
:::

::: {.callout-warning}
## Reproducibility is graded
Before you submit: restart R (`Ctrl/Cmd + Shift + F10`) and render the document
from a clean session. If it does not run start to finish, it is incomplete ---
regardless of how good the maps are. This is the habit that matters most, and
we start enforcing it now.
:::

```{r}
#| label: setup

library(sf)
library(dplyr)
library(ggplot2)
library(tmap)
# add anything else you use -- and load it here, not halfway down

nc <- st_read(system.file("shape/nc.shp", package = "sf"), quiet = TRUE)
```

---

## Part A --- Read and describe the data (2 points)

Answer each in **one or two sentences of prose**, outside the code chunks, and
show the code you used to find out.

1. How many features does `nc` contain, and what does one row represent?
2. What geometry type is stored? Why is it that type rather than `POLYGON`?
   *(Look at the coast.)*
3. What coordinate reference system is the data in? Give its name, say whether it
   is geographic or projected, and say how you found out.
4. In your own words, what do `BIR74` and `SID74` measure? Why does having both
   matter more than having either one alone?

##Answer
1. NC contains 100 features and one row represent  one NC county.
2. Geometry type: Multipolygon because some counties have separate polygon. it is also mentionable that county boundary has  separated coastal areas or island areas.
3.The 'nc' dataset uses the NAD27 geographic coordinate reference system (EPSG:4267), meaning its coordinates represent longitude and latitude rather than projected meters. I checked this with st_crs(nc), st_crs(nc)$input to check the exact EPSG code, and st_is_longlat(nc) to confirm it relies on geographic coordinates.
4. BIR74 shows the number of live births from 1974 to 1978, while SID74 shows the number of SIDS deaths during the same period. Both are needed because counties with more births may naturally have more deaths. Using the two variables together lets us calculate the SIDS death rate per 1,000 live births and compare counties more fairly
### Starting code

```{r}
#| label: a-starter

# A1 -- how big is it, and what are the columns?
nrow(nc)
names(nc)
nc                       # the header block above the table says a lot

# A2 -- one geometry type per feature; count them
table(st_geometry_type(nc))

# A3 -- three ways of asking about the CRS
st_crs(nc)               # the full definition
st_crs(nc)$input         # the short name
st_is_longlat(nc)        # TRUE = geographic (degrees), FALSE = projected (metres)

# A4 -- the help page documents every column
?nc
```

---

## Part B --- Coordinate reference systems (2 points)

1. Transform `nc` to **EPSG:32119** (NAD83 / North Carolina State Plane, in
   metres) and store it as a **new object**. Do not overwrite `nc` --- you need
   both.
2. Plot the counties **before and after**, near enough to each other that a reader
   can compare them. One or two sentences on what visibly changed.
3. Compute the area of every county in **square kilometres**, using the projected
   version. Report the five largest.
4. Now compute areas from the *unprojected* object as well. Compare the two sets
   of numbers and explain, in two or three sentences, what `sf` did and why the
   projected version is the one you would report in a paper.
   
##Answer
1. stored as nc_proj and leaving the original geographic object nc unchanged.
2. The two maps show the same general pattern of counties. The projected map mainly changes the coordinate system, using metres to measure location and distance. The shapes of the counties look almost the same, so the main difference is how their positions and distances are measured.
3.The five largest counties based on the projected area calculation are Robeson, Sampson, Columbus, Bladen, and wake. The code below shows their exact area in square kilometres using the EPSG:32119 coordinate system
4. When st_area() is used on geographic data, it calculates area based on the Earth’s curved surface. Because of this, the results are close to, but slightly different from, the areas calculated using a projected coordinate system. I would use the projected values because EPSG:32119 measures distances in metres and is specifically designed for North Carolina, making it more suitable for calculating county areas in this analysis

::: {.callout-note}
Step 4 will not give you an error, and the two sets of numbers will be close but
not identical. That is the interesting part --- the question is which *surface*
each answer refers to.
:::

### Starting code

```{r}
#| label: b-starter

# B1
nc_proj <- st_transform(nc, 32119)

# B2 -- one plot each
p1 <- ggplot(nc)      + geom_sf(fill = "grey90", colour = "white") + theme_void()
p2 <- ggplot(nc_proj) + geom_sf(fill = "grey90", colour = "white") + theme_void()

p1
p2
# side by side, if you like:  library(patchwork);  p1 + p2

# B3 -- st_area() returns square METRES, as a units object
nc_proj <- nc_proj |>
  mutate(area_km2 = as.numeric(units::set_units(st_area(nc_proj), km^2)))

nc_proj |>
  st_drop_geometry() |>
  slice_max(area_km2, n = 5) |>
  select(NAME, area_km2)

# B4 -- now do the same starting from nc (degrees) and compare

nc_geo <- nc |>
  mutate(area_km2_geo = as.numeric(units::set_units(st_area(nc), km^2)))

area_compare <- nc_proj |>
  st_drop_geometry() |>
  select(NAME, area_km2) |>
  left_join(
    nc_geo |>
      st_drop_geometry() |>
      select(NAME, area_km2_geo),
    by = "NAME"
  ) |>
  mutate(difference_km2 = area_km2_geo - area_km2)

area_compare |>
  arrange(desc(area_km2)) |>
  slice_head(n = 5)
```


## Part C --- Choropleth maps (5 points)

This is the largest part of the exercise, and it is as much about judgement as
about code.

1. Add a column giving the **SIDS rate per 1,000 live births** for 1974--78.
2. Make a choropleth of the **raw counts** (`SID74`) and one of the **rate**,
   using `ggplot2`. Then write three or four sentences explaining why the two maps
   disagree, and which one you would put in front of a health department, and why.
3. Rebuild the rate map using **`tmap`**. Same information, other package.
4. Vary the **class intervals** on the rate map: produce it at least three ways
   --- for instance equal intervals, quantiles, and Jenks natural breaks. Two or
   three sentences on how much the apparent geography of risk changes. Name the
   classification you would defend, and say why.
5. Every map you submit must have a **legend with units**, a **title**, thin
   boundaries that do not overwhelm the fill, and a **perceptually uniform,
   colour-blind-safe palette**. State in one sentence which palette you chose and
   why, including whether dark means high or low.
   
##Answer
1. Column is added. Need to check C1
2.TThe raw-count map and rate map can disagree because counties have very different numbers of live births. A county with more births may have more SIDS deaths simply because more babies are born there, while a smaller county may have fewer deaths but a higher SIDS rate. For a health department, the rate map is more useful because it adjusts the number of deaths based on the number of live births, making comparisons between counties fairer.
3.Map Plotted
4.Equal intervals, quantiles, and Jenks breaks can produce different map patterns because they group the SIDS rates in different ways. Quantiles place a similar number of counties in each of the five groups, while equal intervals may result in some groups having very few counties. Jenks breaks create groups based on natural gaps in the data. I would use *quantiles* here because the main purpose is to compare relative risk across counties. Quantiles make it easier to identify which counties are in the higher-risk groups without allowing a few unusually high values to control the overall map pattern.
5.I chose the viridis palette because it is perceptually uniform and colour-blind-safe, and with direction = -1, darker colours represent higher SIDS values or rates, while lighter colours represent lower values.

::: {.callout-tip}
## What earns full marks here
Not the prettiest map --- the map whose choices you can defend. A student who
picks quantiles and explains the trade-off in two clear sentences scores higher
than one who picks Jenks because it looked nicer.
:::

### Starting code

```{r}
#| label: c-starter

# C1 -- deaths per 1,000 live births
nc <- nc |>
  mutate(rate74 = SID74 / BIR74 * 1000)

# C2a -- raw counts
ggplot(nc) +
  geom_sf(aes(fill = SID74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(
    option = "viridis", direction = -1,
    name = "SIDS deaths"
  ) +
  labs(title = "SIDS deaths, 1974–78") +
  theme_void()

# C2b -- rates
ggplot(nc) +
  geom_sf(aes(fill = rate74), colour = "white", linewidth = 0.15) +
  scale_fill_viridis_c(
    option = "viridis", direction = -1,
    name = "Deaths per 1,000 live births"
  ) +
  labs(title = "SIDS rate, 1974–78") +
  theme_void()

# other palettes to try:
#   scale_fill_viridis_c(option = "viridis" / "plasma" / "cividis" / "mako")
#   scale_fill_distiller(palette = "YlOrRd", direction = 1)
#   RColorBrewer::display.brewer.all()   # shows every ColorBrewer option

# C3 -- the same map in tmap
tmap_mode("plot")                        # "view" makes it interactive

tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.legend = tm_legend(title = "Deaths per 1,000 live births")
  ) +
  tm_title("SIDS rate, 1974–78")

# C4 -- change ONLY the classification, so the comparison is fair
tm_shape(nc) +
  tm_polygons(fill = "rate74",
              fill.scale = tm_scale_intervals(style = "quantile", n = 5)) +
  tm_title("Quantiles")

# Equal intervals
tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.scale = tm_scale_intervals(style = "equal", n = 5)
  ) +
  tm_title("SIDS rate — Equal intervals")

# Jenks natural breaks
tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.scale = tm_scale_intervals(style = "jenks", n = 5)
  ) +
  tm_title("SIDS rate — Jenks natural breaks")

# Quantiles (the supplied comparison)
tm_shape(nc) +
  tm_polygons(
    fill = "rate74",
    fill.scale = tm_scale_intervals(style = "quantile", n = 5)
  ) +
  tm_title("SIDS rate — Quantiles")
```

::: {.callout-warning}
## A note on `tmap` versions
`tmap` changed its argument names in version 4. If an example you find online
uses `tm_polygons(col = ...)` or `style = ...` directly and yours errors, check
`packageVersion("tmap")` and read the help page rather than guessing. The code
above is written for version 4.
:::

---

## Part D --- A point layer (2 points)

1. Look up the approximate longitude and latitude of **four cities in North
   Carolina** and build a small table holding the name and the two coordinates.
   Cite where you got the coordinates.
2. Turn that table into an `sf` object with the correct CRS. State which EPSG
   code you used and why.
3. Plot the four cities **on top of** your rate choropleth, labelled legibly.
4. One sentence: what would go wrong if you built the `sf` object with the
   coordinate columns in the other order?

##Answer
1. I used Durham, Fayetteville, Wilmington, and Winston-Salem as the four city points. Their approximate longitude and latitude coordinates were obtained from the U.S. Census Bureau TIGERweb, State of North Carolina Incorporated Places data.
2.I created the city points using EPSG:4269 (NAD83) because the coordinates are longitude and latitude from the U.S. Census Bureau, then transformed them to EPSG:4267 (NAD27) to match the CRS of the nc county layer.
3. Plotted and labelled
4.If the coordinate columns were reversed, R would treat latitude as longitude and longitude as latitude, so the city points would be plotted in the wrong locations, likely far outside North Carolina.


### Starting code

```{r}
#| label: d-starter

# D1 -- approximate city coordinates from
# U.S. Census Bureau TIGERweb, North Carolina Incorporated Places

cities <- tibble(
  name = c("Durham", "Fayetteville", "Wilmington", "Winston-Salem"),
  lon  = c(-78.8981, -78.9732, -77.8858, -80.2608),
  lat  = c(35.9779, 35.0816, 34.2092, 36.1029)
)
cities

# D2 -- Census TIGER coordinates use NAD83
# EPSG:4269 = NAD83 geographic longitude/latitude

cities_sf <- cities |>
  st_as_sf(
    coords = c("lon", "lat"),
    crs = 4269
  )

# Transform the city points to the CRS used by the nc county layer
cities_sf <- st_transform(cities_sf, st_crs(nc))

# Check that both layers now have the same CRS
st_crs(cities_sf)
st_crs(nc)

# D3 -- plot cities on top of the SIDS rate choropleth

ggplot() +
  geom_sf(
    data = nc,
    aes(fill = rate74),
    colour = "white",
    linewidth = 0.15
  ) +
  geom_sf(
    data = cities_sf,
    size = 2.5,
    colour = "black"
  ) +
  geom_sf_text(
    data = cities_sf,
    aes(label = name),
    size = 3,
    nudge_y = 0.15
  ) +
  scale_fill_viridis_c(
    option = "viridis",
    direction = -1,
    name = "Deaths per 1,000\nlive births"
  ) +
  labs(
    title = "SIDS Rate and Selected North Carolina Cities, 1974–78"
  ) +
  theme_void()
```

---

## Part E --- Reproducibility and AI use (1 point)

1. Your document renders from a clean session, with every package loaded
   explicitly. *(Check this last, after you have stopped editing.)*
2. End the document with an **AI use note**: which tools you used, what for, and
   at which stage --- planning, coding, debugging, interpreting, or writing. Two
   or three sentences.
   
#Answer

2. I used ChatGPT to help with parts of the coding and checking errors. I reviewed the code and explanations before including them in the final assignment.

Remember the course policy: AI use is unrestricted and costs you nothing, but you
must be able to explain every line you hand in. Disclosure is the only
requirement.

### Starting code

```{r}
#| label: e-starter

sessionInfo()
```

---

## Rubric

| Part | What is assessed | Points |
|---|---|---|
| **A** | Correct inspection of the object; clear, accurate answers | 2 |
| **B** | Correct transformation and areas; comparison explained | 2 |
| **C** | Both packages working; counts-vs-rates argument; class intervals compared and a defended choice; cartographic standards met | 5 |
| **D** | Point layer built with the right CRS and correctly overlaid | 2 |
| **E** | Renders from a clean session; AI use note present | 1 |
| | **Total** | **12** |

Writing counts for roughly 10% of the mark within each part: the prose should be
in complete sentences and free of obvious errors. Every part asks for
interpretation as well as code --- a submission that is all code and no sentences
cannot score above about 60%.

---

## Getting help

- Office hours, or email me --- with the code and the exact error message, not a
  screenshot of the whole console.
- The `sf` documentation is genuinely good: `vignette("sf1", package = "sf")`.
- Moraga Chapter 2 covers most of Parts A--C.
- If you are stuck for more than about thirty minutes on something that feels
  mechanical, that is a signal to ask rather than to persevere.